---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggfx

<!-- badges: start -->
[![R build status](https://github.com/thomasp85/ggfx/workflows/R-CMD-check/badge.svg)](https://github.com/thomasp85/ggfx/actions?workflow=R-CMD-check)
[![CRAN status](https://www.r-pkg.org/badges/version/ggfx)](https://CRAN.R-project.org/package=ggfx)
[![CRAN_Download_Badge](http://cranlogs.r-pkg.org/badges/ggfx)](https://CRAN.R-project.org/package=ggfx)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

ggfx is a (currently experimantal) package that allows the use of various 
filters and shaders on ggplot2 layers. At the moment it implements a blur filter
but more will be added, along with the possibility to use custom filters.

## Installation
ggfx is not on CRAN yet (but will probably at some point). In the mean time, if
you are craving for that sweet sweet blur, you can install it from github:

``` r
# install.packages('devtools')
devtools::install_github('thomasp85/ggfx')
```

## Example
The basic API of ggfx is to provide a range of `with_*()` modifier functions 
instead of special versions of common geoms. This means that ggfx will work with
any geom from ggplot2 and the extension packages (I think...). And example of
blurring a point geom can be seen below.

```{r example, message=FALSE}
library(ggplot2)
library(ggfx)
ggplot(mtcars, aes(mpg, disp)) + 
  with_blur(geom_point(size = 3), sigma = 3) + 
  geom_point()
```

## Code of Conduct
Please note that the ggfx project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/1/0/0/). By contributing to this project, you agree to abide by its terms.
